Manuscript title here

### 2024-09-20 manuscript.Rmd
Version 2.2 för manus och 2.0 för projekt redogörelse.

1  Varför trauma är viktigt
    Mortalitet för unga
    Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
      Kanske att chock är stor del av mortalitet.

2  Definiton av chock
    olika typer av chock: hypovolem, obstruktiv, neurogen, kardiogen och distributiv.
    trauma oftas de 3 första??

3  Klassifikationer av chock
    ATLS klassifikation för blödningschock.
      Hur ska jag ta med:
        Obstruktiv som tamponad?
        Neurogen som trauma mot halsrygg?
      RETTS och NEWS istället?? de används vid triage men är inte chock specifik.
    Kanske borde använda ATLS 4 kategorier + obstruktiv + neurogen vid sidan??

  (4  Kliniskt för att förbättra omhändertagande används OFI
    Förklara vad OFI är för något
      To be able to assess and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI). OFI is multidiciplinary morbidity and mortality reviews, where opportunity for improvements are identified in the structure and clinical process of trauma care[@albaaj_patient_2023]. )

5  Frågeställning:
    a, Vilka typer av OFI finns hos chockpatienter?
    b, Koppling mellan nivå av chock och sannolikhet till OFI.
    Vilken av de två är huvudfrågeställningen?
    PICO: Population: Chock pat, bland 8000 (2014-2023)
          Intervention: OFI efter vård?
          Comparison: mellan chocknivåer?
          Outcome: Sannolikhet till OFI?
      Det blir ju inte direkt någon ja/nej svar på frågeställningen...funkar det?

Trauma is a major global public health concern. It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate]. Shock is a major cause of mortality in trauma patients. The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023]. There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive. These causes can occur alone or in combination. The most usual one during trauma are hypovolemic in form of a major haemorrhage [@teixeira_preventable_2007].

Clinically to be able to differentiate the degree of shock in context of trauma, the ATLS classification for major haemorrhage can be used. The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only suits hypovolemic shocks and not the other types. (Källa) There is other classification for triaging or tracking the clinical deterioration of patients, which can be used for other causes of shocks. The common denominator for these classifications is they are all based on vital parameters.

The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010]. To improve trauma care quality, healthcare systems identify Opportunities for Improvement (OFI) through structured reviews of patient cases. Multidisciplinary morbidity and mortality reviews play a central role in this process, where clinicians from different specialities together analyse trauma cases to evaluate both the structure and clinical processes involved in trauma care following the Donabedian’s quality of care framework[@albaaj_patient_2023]. For shock patients, timely interventions such as fluid resuscitation/blood transfusion, intervention, OR and radiology are critical, and delays or missteps can result in adverse outcomes (källa). By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity.

>>>>

Trauma is a major global public health concern. It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate]. The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010] [@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury[@berry_shock_2015]. It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not[@vang_shock_2022].

The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023]. There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive. These causes can occur alone or in combination. The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].

To differentiate the severity of trauma shock, we can divide shock into different degrees. One of such classification is the ATLS classification for major haemorrhage. The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classification for predicting massive transfusion (TASH, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).

One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI). This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals. There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [källa 8]. In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988]. It is based on the three factors of structure, process and outcome. By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?]. This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].

### 2024-09-20 project-plan.Rmd
Version 2.2 för manus och 2.0 för projekt redogörelse.

---
title: "Manuscript title here"
subtitle: "Add a Subtitle if Needed"
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
---

@@

<!-- The project plan should be 3-4 pages long and written in
Rmarkdown (like this document). Remove all comments in italic when you 
use this document as a template for your own project plan. -->

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
the console and you should only see headings -->

Introduction
============

*This is part A) in the project plan [sv: projektredogörelse]. *

@@

*bibliography.bib includes an example reference. Add additional
references in that file as bibtex entries and then cite in text using:
"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
https://www.doi2bib.org/. If you use Zotero you can simply export 
entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*

Trauma is a significant global public health concern, responsible for causing death and disability in over four million people each year. Among those affected, younger populations are disproportionately impacted, making trauma management a critical issue in healthcare systems worldwide[@noauthor_injuries_nodate]. The initial management of trauma patients in the hospital setting is extremely time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010]. Trauma patients presenting in shock have a notably higher mortality rate compared to those who are not in shock, underscoring the need for improved care strategies for these high-risk patients[@vang_shock_2022].
Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023]. There are four primary causes of shock: hypovolemic, obstructive, cardiogenic, and distributive. These etiologies can occur in isolation or in combination. In the context of trauma, hypovolemic shock, usually caused by massive hemorrhage, is the most common type of shock[@teixeira_preventable_2007].

To classify shock in trauma 
The Advanced Trauma Life Support (ATLS) system provides a widely used classification for hypovolemic shock based on blood loss, heart rate, blood pressure, respiratory rate, mental status, and urine output. ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe, and helps guide resuscitative efforts. However, this classification is specific to hypovolemic shock and does not account for the other types of shock.

To address the broader spectrum of shock types encountered in trauma, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are commonly used in emergency care settings. While these systems are not shock-specific, they offer useful frameworks for assessing patients in various shock states by prioritizing vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR), and mental status.

RETTS is designed for rapid triage and treatment prioritization in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency. It incorporates physiological parameters similar to those used in ATLS, making it applicable to patients in various forms of shock, including hypovolemic, cardiogenic, and distributive shock.

NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of hospitalized patients. Like ATLS and RETTS, it evaluates HR, BP, RR, and mental status, but also includes oxygen saturation and temperature. NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock.

Despite their differences, all three systems—ATLS, RETTS, and NEWS2—share key similarities, particularly in their use of vital sign markers such as heart rate, blood pressure, respiratory rate, and mental status. This overlap suggests a potential framework for integrating these systems to better classify trauma patients presenting in any form of shock.

Infoga OFI

@@

Methods
=======

@@

Project Update and Timeline
===========================

*This is part B) in the project plan [sv: projektredogörelse]. *

@@

Back-up Plan
============

@@

References
==========

>>>>

---
title: "Manuscript title here"
subtitle: "Add a Subtitle if Needed"
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{=html}
<!-- The project plan should be 3-4 pages long and written in
Rmarkdown (like this document). Remove all comments in italic when you 
use this document as a template for your own project plan. -->
```
```{=html}
<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->
```
```{=html}
<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
the console and you should only see headings -->
```
# Introduction

*This is part A) in the project plan [sv: projektredogörelse].*

@@

*bibliography.bib includes an example reference. Add additional
references in that file as bibtex entries and then cite in text using:
"[@exampleKey9999]". One can convert DOIs' to bibtex entries using
<https://www.doi2bib.org/>. If you use Zotero you can simply export
entries to bibtex `Ctrl-Shift-C`, however you need to first setup Quick
Copy to export to bibtex
(<https://www.zotero.org/support/creating_bibliographies#quick_copy>)*

Trauma is a major global public health concern. It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate]. The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010] [@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury[@berry_shock_2015]. It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not[@vang_shock_2022].

The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023]. There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive. These causes can occur alone or in combination. The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].

To differentiate the severity of trauma shock, we can divide shock into different degrees. One of such classification is the ATLS classification for major haemorrhage. The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classification for predicting massive transfusion (TASH, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).

One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI). This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals. There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [källa 8]. In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988]. It is based on the three factors of structure, process and outcome. By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?]. This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].

@@

# Methods

@@

# Project Update and Timeline

*This is part B) in the project plan [sv: projektredogörelse].*

@@

# Back-up Plan

@@

# References
